---
title: "Biomass bins and age structure"
subtitle: "Summer flounder, scup, and black sea bass"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_fold: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Create a range of biomass and associated age structures for summer flounder based on the 2023 MT assessment. This will be used to parameterize a relationship between summer flounder population structure and expectation of catch given regs from the recreational demand model.

## Approach

Take the 2023 MT assessment input and output files as a starting point. Find max and min B relative to the reference point and most extreme age structures (all young, all old) from the assessment. Can we get a relationship (or eliminate unlikely combinations) of B/Bmsy and age structure that is realistic for this stock?

## Range of B/Bmsy from assessment

### Summer flounder

Mark Terciero provided the files from the 2023 MT assessment. They are in this repo in the `fluke2023MTfiles` folder: https://github.com/sgaichas/MAFMCrecMeasures/tree/main/fluke2023MTfiles

First, here in mac land we need to change filenames to have lowercase extensions. Done by hand.

We want the following from the .rep file: 

+  Population Numbers at the Start of the Year 
+  Biomass Time Series

This is findable in the .rdat file using Chris Legault and Liz Brook's ASAPplots functions:

```{r}
# try Legault's asapPlots functions to read rep/rdat

#devtools::install_github("cmlegault/ASAPplots", build_vignettes = TRUE)

library(ASAPplots)

asap.name <- "F2023_NEWSET_ESS_V1"    
wd <- here::here("fluke2023MTfiles")

#test from plotASAP.r lines 42-43
rdat <- paste0(asap.name,".rdat")
asap <- dget(paste0(wd,"/",rdat)) # modified from windows specific backslash

# # desired objects
# asap$SSB
# 
# asap$exploitable.B
# 
# asap$tot.jan1.B
# 
# asap$N.age


```

Add reference points from assessment: 

Download from https://apps-nefsc.fisheries.noaa.gov/saw/sasi.php

Search Criteria Used

+  Assessment Year:	2023
+  Species Name:	Summer flounder
+  Stock Area:	Summer flounder
+  Review Type:	Level 2 Expedited Review
+  Information Type:	All

Table 2 of `Summer_flounder_MTA_2023_06_08_130636.678006.pdf` has reference points

$SSB_{msy}$ in 2023 is 49561

Show SSB from assessment relative to $SSB_{msy}$:

```{r}

SSBmsy <- 49561

relSSB <- asap$SSB/SSBmsy

years <- asap$parms$styr:asap$parms$endyr

SSBdat <- data.frame(years = years,
                     relSSB = relSSB,
                     SSB = asap$SSB,
                     explB = asap$exploitable.B,
                     totB = asap$tot.jan1.B)

ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSB)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw() +
  ggplot2::ggtitle("Fluke")
```

Lines represent three "bins" with B > Bmsy in the top bin, 0.5 Bmsy <B < B msy in the center and B < 0.5 Bmsy at the bottom.

B has never been above 1.5 Bmsy in this time series. Will have to punt on that one or use max B.


Summarize in each bin:
```{r}
SSBbins <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSB > 1.5 ~ "high above target",
                                dplyr::between(relSSB, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSB, 0.5,1.0) ~ "below target",
                                relSSB < 0.5 ~ "overfished"))

binsummary <- SSBbins |> 
  dplyr::group_by(bin) |>
  dplyr::summarise(meanSSB = mean(SSB),
                   meanexplB = mean(explB),
                   meantotB = mean(totB),
                   nyears = dplyr::n())

binsummary
```

We could characterize age structure associated with B in each bin. 

```{r}

Nage <- asap$N.age |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbins |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSB, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```

### Scup

Mark Terciero provided the files from the 2023 MT assessment. They are in this repo in the `scup2023MTfiles` folder: https://github.com/sgaichas/MAFMCrecMeasures/tree/main/scup2023MTfiles

First, here in mac land we need to change filenames to have lowercase extensions. Done by hand.

We want the following from the .rep file: 

+  Population Numbers at the Start of the Year 
+  Biomass Time Series

This is findable in the .rdat file using Chris Legault and Liz Brook's ASAPplots functions:

```{r}
# try Legault's asapPlots functions to read rep/rdat

#devtools::install_github("cmlegault/ASAPplots", build_vignettes = TRUE)

library(ASAPplots)

asap.name <- "MTA2023_NEWSET_ESS_V1"    
wd <- here::here("scup2023MTfiles")

#test from plotASAP.r lines 42-43
rdat <- paste0(asap.name,".rdat")
asap <- dget(paste0(wd,"/",rdat)) # modified from windows specific backslash

# desired objects
# asap$SSB
# 
# asap$exploitable.B
# 
# asap$tot.jan1.B
# 
# asap$N.age


```

Add reference points from assessment: 

Download from https://apps-nefsc.fisheries.noaa.gov/saw/sasi.php

Search Criteria Used

+  Assessment Year:	2023
+  Species Name:	Scup
+  Stock Area:	UNIT
+  Review Type:	Level 2 Expedited Review
+  Information Type:	All

Table 2 of `Scup_2023_MTA_2023_06_05_104930.21231.pdf` has reference points

$SSB_{msy}$ in 2023 is 78593

Show SSB from assessment relative to $SSB_{msy}$:

```{r}

SSBmsy <- 78593

relSSB <- asap$SSB/SSBmsy

years <- asap$parms$styr:asap$parms$endyr

SSBdat <- data.frame(years = years,
                     relSSB = relSSB,
                     SSB = asap$SSB,
                     explB = asap$exploitable.B,
                     totB = asap$tot.jan1.B)

ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSB)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw() +
  ggplot2::ggtitle("Scup")
```

Lines represent three "bins" with B > Bmsy in the top bin, 0.5 Bmsy <B < B msy in the center and B < 0.5 Bmsy at the bottom.

B has been above 1.5 Bmsy in this time series so we can summarize age comps in all 4 bins.

Summarize in each bin:
```{r}
SSBbins <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSB > 1.5 ~ "high above target",
                                dplyr::between(relSSB, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSB, 0.5,1.0) ~ "below target",
                                relSSB < 0.5 ~ "overfished"))

binsummary <- SSBbins |> 
  dplyr::group_by(bin) |>
  dplyr::summarise(meanSSB = mean(SSB),
                   meanexplB = mean(explB),
                   meantotB = mean(totB),
                   nyears = dplyr::n())

binsummary
```

We now characterize age structure associated with B in each bin. 

```{r}

Nage <- asap$N.age |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbins |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSB, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```

### Black Sea Bass

Kiersten Curti provided a link to the preliminary (pre-review but public) 2023 RT results: https://github.com/kcurti/BSB.2023.RT.Modeling/tree/main/2023.RT.Runs/Run34 

We want the following from the output files: 

+  Population Numbers at the Start of the Year 
+  Biomass Time Series

Black Sea Bass is assessed in two areas, North and South, but reference points and status apply to the entire population. I'll do my best to bin things to full population but also give North and South results if that is useful for the rec demand model.

Explore the WHAM rdata file:

```{r}
# rdata file from repo

ghrdat <- "https://github.com/kcurti/BSB.2023.RT.Modeling/raw/main/2023.RT.Runs/Run34/outputs/Run34.WHAM.Outputs.RDATA"

download.file(ghrdat, here::here("BSB2023RTfiles/Run34.WHAM.Outputs.Rdata"))
load("BSB2023RTfiles/Run34.WHAM.Outputs.Rdata")

#objects of interest

# SSB.yr.north
# 
# NAA.north
# 
# SSB.yr.south
# 
# NAA.south


```

Reference points from WHAM output `brps`

This plot is for the whole population (north plus south)

```{r}
SSBmsynorth <- brps$est[2]
SSBmsysouth <- brps$est[3]

SSBmsy <- brps$est[4]

relSSBnorth <- SSB.yr.north$est/SSBmsynorth
relSSBsouth <- SSB.yr.south$est/SSBmsysouth

relSSBtot <- (SSB.yr.north$est + SSB.yr.south$est)/SSBmsy

years <- model.yrs

SSBdat <- data.frame(years = years,
                     relSSB = relSSBtot,
                     SSB = SSB.yr.north$est + SSB.yr.south$est,
                     SSBnorth = SSB.yr.north$est,
                     SSBsouth = SSB.yr.south$est,
                     relSSBnorth = relSSBnorth,
                     relSSBsouth = relSSBsouth)

ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSB)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw() +
  ggplot2::ggtitle("BSB")
```

And here it is for each area. Both areas and the full population have been in all 4 bins.

```{r}

north <- ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSBnorth)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw() +
  ggplot2::ggtitle("North BSB")

south <- ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSBsouth)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw() +
  ggplot2::ggtitle("South BSB")

library(patchwork)

north + south

```

Summarize in each bin (full population):
```{r}
SSBbins <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSB > 1.5 ~ "high above target",
                                dplyr::between(relSSB, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSB, 0.5,1.0) ~ "below target",
                                relSSB < 0.5 ~ "overfished"))

binsummary <- SSBbins |> 
  dplyr::group_by(bin) |>
  dplyr::summarise(meanSSB = mean(SSB),
                   nyears = dplyr::n()
                   )

binsummary

binsummaryNorth <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSBnorth > 1.5 ~ "high above target",
                                dplyr::between(relSSBnorth, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSBnorth, 0.5,1.0) ~ "below target",
                                relSSBnorth < 0.5 ~ "overfished")) |>
   dplyr::group_by(bin) |>
   dplyr::summarise(meanSSBnorth = mean(SSBnorth),
                   nyears = dplyr::n()
                   )
 

binsummarySouth <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSBsouth > 1.5 ~ "high above target",
                                dplyr::between(relSSBsouth, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSBsouth, 0.5,1.0) ~ "below target",
                                relSSBsouth < 0.5 ~ "overfished")) |>
   dplyr::group_by(bin) |>
   dplyr::summarise(meanSSBsouth = mean(SSBsouth),
                   nyears = dplyr::n()
                   )

```

Summary for north:
```{r}
binsummaryNorth
```

Summary for south:
```{r}
binsummarySouth
```

Age structure, full population: 
```{r}

NAAtot <- NAA.north + NAA.south

Nage <- NAAtot |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbins |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSB, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```

Age structure, North: 
```{r}

SSBbinsN <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSBnorth > 1.5 ~ "high above target",
                                dplyr::between(relSSBnorth, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSBnorth, 0.5,1.0) ~ "below target",
                                relSSBnorth < 0.5 ~ "overfished"))

Nage <- NAA.north |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbinsN |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSBnorth, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```

Age structure, South:

```{r}
SSBbinsS <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSBsouth > 1.5 ~ "high above target",
                                dplyr::between(relSSBsouth, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSBsouth, 0.5,1.0) ~ "below target",
                                relSSBsouth < 0.5 ~ "overfished"))

Nage <- NAA.south |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbinsS |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSBsouth, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```