---
title: "Biomass bins and age structure"
subtitle: "Summer flounder, scup, and black sea bass"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_fold: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Create a range of biomass and associated age structures for summer flounder based on the 2023 MT assessment. This will be used to parameterize a relationship between summer flounder population structure and expectation of catch given regs from the recreational demand model.

## Approach

Take the 2023 MT assessment input and output files as a starting point. Find max and min B relative to the reference point and most extreme age structures (all young, all old) from the assessment. Can we get a relationship (or eliminate unlikely combinations) of B/Bmsy and age structure that is realistic for this stock?

## Range of B/Bmsy from assessment

### Summer flounder

Mark Terciero provided the files from the 2023 MT assessment. They are in this repo in the `fluke2023MTfiles` folder: https://github.com/sgaichas/MAFMCrecMeasures/tree/main/fluke2023MTfiles

First, here in mac land we need to change filenames to have lowercase extensions. Done by hand.

We want the following from the .rep file: 

+  Population Numbers at the Start of the Year 
+  Biomass Time Series

This is findable in the .rdat file using Chris Legault and Liz Brook's ASAPplots functions:

```{r}
# try Legault's asapPlots functions to read rep/rdat

#devtools::install_github("cmlegault/ASAPplots", build_vignettes = TRUE)

library(ASAPplots)

asap.name <- "F2023_NEWSET_ESS_V1"    
wd <- here::here("fluke2023MTfiles")

#test from plotASAP.r lines 42-43
rdat <- paste0(asap.name,".rdat")
asap <- dget(paste0(wd,"/",rdat)) # modified from windows specific backslash

# # desired objects
# asap$SSB
# 
# asap$exploitable.B
# 
# asap$tot.jan1.B
# 
# asap$N.age


```

Add reference points from assessment: 

Download from https://apps-nefsc.fisheries.noaa.gov/saw/sasi.php

Search Criteria Used

+  Assessment Year:	2023
+  Species Name:	Summer flounder
+  Stock Area:	Summer flounder
+  Review Type:	Level 2 Expedited Review
+  Information Type:	All

Table 2 of `Summer_flounder_MTA_2023_06_08_130636.678006.pdf` has reference points

$SSB_{msy}$ in 2023 is 49561

Show SSB from assessment relative to $SSB_{msy}$:

```{r}

SSBmsy <- 49561

relSSB <- asap$SSB/SSBmsy

years <- asap$parms$styr:asap$parms$endyr

SSBdat <- data.frame(years = years,
                     relSSB = relSSB,
                     SSB = asap$SSB,
                     explB = asap$exploitable.B,
                     totB = asap$tot.jan1.B)

ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSB)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw()
```

Lines represent three "bins" with B > Bmsy in the top bin, 0.5 Bmsy <B < B msy in the center and B < 0.5 Bmsy at the bottom.

B has never been above 1.5 Bmsy in this time series. Will have to punt on that one or use max B.


Summarize in each bin:
```{r}
SSBbins <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSB > 1.5 ~ "high above target",
                                dplyr::between(relSSB, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSB, 0.5,1.0) ~ "below target",
                                relSSB < 0.5 ~ "overfished"))

binsummary <- SSBbins |> 
  dplyr::group_by(bin) |>
  dplyr::summarise(meanSSB = mean(SSB),
                   meanexplB = mean(explB),
                   meantotB = mean(totB),
                   nyears = dplyr::n())

binsummary
```

We could characterize age structure associated with B in each bin. 

```{r}

Nage <- asap$N.age |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbins |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSB, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```

### Scup

Mark Terciero provided the files from the 2023 MT assessment. They are in this repo in the `scup2023MTfiles` folder: https://github.com/sgaichas/MAFMCrecMeasures/tree/main/scup2023MTfiles

First, here in mac land we need to change filenames to have lowercase extensions. Done by hand.

We want the following from the .rep file: 

+  Population Numbers at the Start of the Year 
+  Biomass Time Series

This is findable in the .rdat file using Chris Legault and Liz Brook's ASAPplots functions:

```{r}
# try Legault's asapPlots functions to read rep/rdat

#devtools::install_github("cmlegault/ASAPplots", build_vignettes = TRUE)

library(ASAPplots)

asap.name <- "MTA2023_NEWSET_ESS_V1"    
wd <- here::here("scup2023MTfiles")

#test from plotASAP.r lines 42-43
rdat <- paste0(asap.name,".rdat")
asap <- dget(paste0(wd,"/",rdat)) # modified from windows specific backslash

# desired objects
# asap$SSB
# 
# asap$exploitable.B
# 
# asap$tot.jan1.B
# 
# asap$N.age


```

Add reference points from assessment: 

Download from https://apps-nefsc.fisheries.noaa.gov/saw/sasi.php

Search Criteria Used

+  Assessment Year:	2023
+  Species Name:	Scup
+  Stock Area:	UNIT
+  Review Type:	Level 2 Expedited Review
+  Information Type:	All

Table 2 of `Scup_2023_MTA_2023_06_05_104930.21231.pdf` has reference points

$SSB_{msy}$ in 2023 is 78593

Show SSB from assessment relative to $SSB_{msy}$:

```{r}

SSBmsy <- 78593

relSSB <- asap$SSB/SSBmsy

years <- asap$parms$styr:asap$parms$endyr

SSBdat <- data.frame(years = years,
                     relSSB = relSSB,
                     SSB = asap$SSB,
                     explB = asap$exploitable.B,
                     totB = asap$tot.jan1.B)

ggplot2::ggplot(SSBdat, ggplot2::aes(x = years, y = relSSB)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 1.5) + 
  ggplot2::geom_hline(yintercept = 1) + 
  ggplot2::geom_hline(yintercept = 0.5) + 
  ggplot2::theme_bw()
```

Lines represent three "bins" with B > Bmsy in the top bin, 0.5 Bmsy <B < B msy in the center and B < 0.5 Bmsy at the bottom.

B has never been above 1.5 Bmsy in this time series. Will have to punt on that one or use max B.


Summarize in each bin:
```{r}
SSBbins <- SSBdat |>
  dplyr::mutate(bin = dplyr::case_when(relSSB > 1.5 ~ "high above target",
                                dplyr::between(relSSB, 1.0,1.5) ~ "above target",
                                dplyr::between(relSSB, 0.5,1.0) ~ "below target",
                                relSSB < 0.5 ~ "overfished"))

binsummary <- SSBbins |> 
  dplyr::group_by(bin) |>
  dplyr::summarise(meanSSB = mean(SSB),
                   meanexplB = mean(explB),
                   meantotB = mean(totB),
                   nyears = dplyr::n())

binsummary
```

We could characterize age structure associated with B in each bin. 

```{r}

Nage <- asap$N.age |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "years") |>
  dplyr::mutate(years = as.integer(years)) |>
  dplyr::rename_with(~ paste0("age", .), -years)

ageSSBbins <- SSBbins |>
  dplyr::left_join(Nage) |>
  dplyr::mutate(total = rowSums(dplyr::across(starts_with("age"))), 
     dplyr::across(starts_with("age"), ~ ./total * 100, .names = 'per_{.col}')) |>
  dplyr::select(years, relSSB, bin, starts_with("per_age")) |>
  tidyr::pivot_longer(starts_with("per_age"), values_to = "per_age", names_to = "age") |>
  tidyr::separate(age, into = c("A", "age"), sep = "per_age") 

ggplot2::ggplot(ageSSBbins, ggplot2::aes(x=age, y=per_age, colour = years)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~bin) +
  ggplot2::theme_bw() +
  ggplot2::ylab("Percent at age")
                  

  
```

